---
title: "KNN decision boundaries"
author: "Eric Vance"
format: pdf
editor: visual
date: 2026-01-20
---

## For Team Application Exercises Wednesday, Jan 21, 2026

I'm going to use the iris dataset to demonstrate fitting knn models and plotting decision boundaries.

## Exercise 2: Plot of iris data

The iris dataset is already built in. Still, I need to install and load some packages for KNN.

```{r}
#install.packages("class")
library(class)
library(tidyverse)
glimpse(iris)
```

Plot Petal.Length (x) v. Petal.width (y).

```{r}
ggplot(data=select(iris,Petal.Length,Petal.Width,Species)) +
  geom_point(aes(x=Petal.Length, y=Petal.Width, color=Species)) +
  labs(title = "3 Species of Iris",
       x = "Petal Length",
       y = "Petal Width")
ggsave("Plots/iris_plot_2026-01-21.pdf")
```

## Exercise 4: Fit KNN model to grid and plot contours (decision boundaries)

We are going to make predictions of which species an iris might be for given values of Petal.Length and Petal.Width. There are 50 training observations that we will train our K-Nearest Neighbors (KNN) model on. Then we'll apply the KNN method to make (test) predictions for the entire 2-dimensional grid of Petal.Width and Petal.Length.

First create a grid.

```{r}
# Define the grid limits
x_min <- min(iris$Petal.Length) - 1
x_max <- max(iris$Petal.Length) + 1
y_min <- min(iris$Petal.Width) - 1
y_max <- max(iris$Petal.Width) + 1

# Create a grid of values by 0.1 x 0.1
grid2 <- expand.grid(Petal.Length = seq(x_min, x_max, by = 0.1),
                    Petal.Width = seq(y_min, y_max, by = 0.1))
```

### Apply KNN

Now we'll fit KNN with k=2.

```{r}
# Train the KNN model and predict on the grid
predicted_species <- knn(train = iris[,3:4], test = grid2, 
                         cl = iris[,5], k = 2)

# Add predictions to the grid data
grid2$Predicted.Species <- predicted_species
```

Now plot the decision boundary by creating a contour plot of the now three dimensional dataset (X1=Petal.Length, X2=Petal.Width, Z=Predicted.Species). The contour plot will be overlaid on the original scatter plot.

```{r}
base.plot <- ggplot(data=select(iris,Petal.Length,Petal.Width,Species)) +
  geom_point(aes(x=Petal.Length, y=Petal.Width, color=Species)) +
  labs(title = "KNN Decision Boundary for k=2",
       x = "Petal Length",
       y = "Petal Width")

# Add decision boundaries using geom_contour with the grid "data"
contour.plot <- base.plot +
  geom_contour(data = grid2, aes(x = Petal.Length, y = Petal.Width, 
                                z = as.numeric(Predicted.Species)),
               bins = 3, color = "black") 
  
print(contour.plot)
ggsave("Plots/iris_KNN.k2_2026-01-21.pdf")
```
