---
title: "Code for 2026-02-25"
format: pdf
editor: visual
---

## Code for TAE

Generating data to determine knots and fit natural and cubic splines.

```{r}
library(splines)
library(tidyverse)
set.seed(225)

x1 <- seq(0, 4 * pi, length.out = 100)
y1 <- sin(x1) + rnorm(100, 0, 0.3)
x2 <- seq(0, 10, length.out = 100)
y2 <- ifelse(x2 < 3, 2 * x2, ifelse(x2 < 7, -x2 + 9, 0.5 * x2 + 2)) + rnorm(100, 0, 0.5)

x <- x1+x2
y <- y1+y2

data <- data.frame(x, y)

ggplot(data, aes(x = x1, y = y1)) +
  geom_point() +
  labs(title = "Nonlinear Data", x = "X", y = "Y") +
  theme_minimal() 
ggsave("2026-02-25.NonlinearData.png")

```

Let's fit all the models to the data and figure out how many knots and then make predictions in the middle using natural spline v. cubic spline.

### Natural Splines

```{r}
library(splines)
library(caret)  # For cross-validation functionality
library(ggplot2) # For plotting

# Sample Data from above

# Define the range of knots to test
knot.counts <- 1:10  # Test from 2 to 10 knots

# Create a data frame to store the results
cv.results <- data.frame(knots = knot.counts, cv.error = NA)

# Perform cross-validation for each number of knots
for (i in 1:length(knot.counts)) {
  num.knots <- knot.counts[i]

  # Define the cross-validation folds
  folds <- createFolds(data$y, k = 10) # 10-fold CV

  # Store the errors for each fold
  fold.errors <- numeric(length(folds))

  # Perform cross-validation
  for (j in 1:length(folds)) {
    train.data <- data[-folds[[j]], ]
    test.data <- data[folds[[j]], ]

    # Fit the natural spline model
    spline.model <- lm(y ~ ns(x, knots = quantile(train.data$x, seq(0, 1, length.out = num.knots + 2))[-c(1, num.knots+2)]), data = train.data) #the knots are the quantiles, but not the first or last (i.e., not the min or max)

    # Make predictions on the test fold
    predictions <- predict(spline.model, newdata = test.data)

    # Calculate the mean squared error (MSE)
    fold.errors[j] <- mean((test.data$y - predictions)^2)
  }

  # Store the average MSE
  cv.results$cv.error[i] <- mean(fold.errors)
}

# Find the optimal number of knots
optimal.knots <- cv.results$knots[which.min(cv.results$cv.error)]
cat("Optimal number of knots:", optimal.knots, "\n")

# Plot the cross-validation results
ggplot(cv.results, aes(x = knots, y = cv.error)) +
  geom_line() +
  geom_point() +
  labs(title = "Cross-Validation for Knot Selection", x = "Number of Knots", y = "Cross-Validation Error (MSE)") +
  theme_minimal()

# Fit the final model with the optimal number of knots
final.spline.model <- lm(y ~ ns(x, knots = quantile(data$x, seq(0, 1, length.out = optimal.knots + 2))[-c(1, optimal.knots+2)]), data = data)

# Visualize the final model
ggplot(data, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(y = predict(final.spline.model)), color = "blue") +
  labs(title = paste("Natural Spline with", optimal.knots, "Knots"), x = "X", y = "Y") +
  theme_minimal()
```

How does this natural spline compare with pre-selected knots at X=11 and X=16?

First, let's redo this analysis with 10x more data.

```{r}
set.seed(225)

x1 <- seq(0, 4 * pi, length.out = 1000)
y1 <- sin(x1) + rnorm(1000, 0, 0.3)
x2 <- seq(0, 10, length.out = 1000)
y2 <- ifelse(x2 < 3, 2 * x2, ifelse(x2 < 7, -x2 + 9, 0.5 * x2 + 2)) + rnorm(1000, 0, 0.5)

x <- x1+x2
y <- y1+y2

data.10x <- data.frame(x, y)

knot.counts <- 1:10  # Test from 2 to 10 knots

# Create a data frame to store the results
cv.results <- data.frame(knots = knot.counts, cv.error = NA)

# Perform cross-validation for each number of knots
for (i in 1:length(knot.counts)) {
  num.knots <- knot.counts[i]

  # Define the cross-validation folds
  folds <- createFolds(data$y, k = 10) # 10-fold CV

  # Store the errors for each fold
  fold.errors <- numeric(length(folds))

  # Perform cross-validation
  for (j in 1:length(folds)) {
    train.data <- data.10x[-folds[[j]], ]
    test.data <- data.10x[folds[[j]], ]

    # Fit the natural spline model
    spline.model <- lm(y ~ ns(x, knots = quantile(train.data$x, seq(0, 1, length.out = num.knots + 2))[-c(1, num.knots+2)]), data = train.data) #the knots are the quantiles, but not the first or last (i.e., not the min or max)

    # Make predictions on the test fold
    predictions <- predict(spline.model, newdata = test.data)

    # Calculate the mean squared error (MSE)
    fold.errors[j] <- mean((test.data$y - predictions)^2)
  }

  # Store the average MSE
  cv.results$cv.error[i] <- mean(fold.errors)
}

# Find the optimal number of knots
optimal.knots.10x <- cv.results$knots[which.min(cv.results$cv.error)]
cat("Optimal number of knots:", optimal.knots, "\n")

# Plot the cross-validation results
ggplot(cv.results, aes(x = knots, y = cv.error)) +
  geom_line() +
  geom_point() +
  labs(title = "Cross-Validation for Knot Selection 10x data", x = "Number of Knots", y = "Cross-Validation Error (MSE)") +
  theme_minimal()

# Fit the final model with the optimal number of knots
final.spline.model.10x <- lm(y ~ ns(x, knots = quantile(data$x, seq(0, 1, length.out = optimal.knots.10x + 2))[-c(1, optimal.knots.10x+2)]), data = data.10x)

# Visualize the final model
ggplot(data.10x, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(y = predict(final.spline.model.10x)), color = "blue") +
  labs(title = paste("Natural Spline with", optimal.knots.10x, "Knots (10x data"), x = "X", y = "Y") +
  theme_minimal()
```

Just two knots actually seems optimal. Those knots would be located at:

```{r}
quantile(x,seq(0,1,length.out=4))# 7.522 and 15.044
```

But still, what is the test MSE for a model with 8, 6, or 2 equally spaced knots compared to knots at x=11 and x=16?

First, let's generate a new validation set of 100 x's using a different seed.

```{r}
set.seed(228)

x1 <- seq(0, 4 * pi, length.out = 100)
y1 <- sin(x1) + rnorm(100, 0, 0.3)
x2 <- seq(0, 10, length.out = 100)
y2 <- ifelse(x2 < 3, 2 * x2, ifelse(x2 < 7, -x2 + 9, 0.5 * x2 + 2)) + rnorm(100, 0, 0.5)

x <- x1+x2
y <- y1+y2
valid.data <- data.frame(x,y)
```

Let's fit the natural spline models (with original data and 10x data) with knots at x=11 and x=16.

```{r}
spline.model.2knots <- lm(y ~ ns(x, knots = c(11,16)), data = data)
spline.model.2knots.10x <- lm(y ~ ns(x, knots = c(11,16)), data = data.10x)
ggplot(valid.data, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(y = predict(spline.model.2knots.10x, newdata=valid.data)), color = "blue") +
  geom_line(aes(y = predict(spline.model.2knots, newdata=valid.data)), color = "red") +
  labs(title = "Natural Spline with 2 pre-specified knots (blue is 10x data)", x = "X", y = "Y") +
  theme_minimal()
```

```{r}
(valid.MSE <- mean((valid.data$y - predict(final.spline.model,newdata=valid.data))^2)) #8 knots [1] 0.4584815
(valid.MSE.10x <- mean((valid.data$y - predict(final.spline.model.10x,newdata=valid.data))^2)) #6 knots [1] 0.4660506

(valid.MSE.2knots <- mean((valid.data$y - predict(spline.model.2knots,newdata=valid.data))^2)) #2 knots, 100 data points #0.867215
(valid.MSE.2knots.10x <- mean((valid.data$y - predict(spline.model.2knots.10x,newdata=valid.data))^2)) #2 knots, 1000 data points #0.8648105

#Same MSE for specifying the knots or not.


```

Same MSE whether I specify the knots or not.

Plot the 2 pre-specified knots vs the 8 knots or 6 knots.

```{r}
ggplot(valid.data, aes(x = x, y = y)) + geom_point() + geom_line(aes(y = predict(spline.model.2knots.10x, newdata=valid.data)), color = "blue") + geom_line(aes(y = predict(final.spline.model.10x, newdata=valid.data)), color = "red") + labs(title = "Natural Spline with 2 pre-specified knots (blue) or 6 knots (red)", x = "X", y = "Y") + theme_minimal()

```

2 pre-specified knots is WAY worse than letting R find the optimal number of knots. And the 8 knot model worked better on the test set than the 6 knot model fit to 1000 data points. More data isn't necessarily better.

### Fit a cubic spline and regression spline (smoothing spline?)

First, let's redo this analysis but adding a cubic spline and a regression spline that allows for discontinuities.

```{r}
set.seed(226)

x1 <- seq(0, 4 * pi, length.out = 100)
y1 <- sin(x1) + rnorm(100, 0, 0.3)
x2 <- seq(0, 10, length.out = 100)
y2 <- ifelse(x2 < 3, 2 * x2, ifelse(x2 < 7, -x2 + 9, 0.5 * x2 + 2)) + rnorm(100, 0, 0.5)

x <- x1+x2
y <- y1+y2

data.test <- data.frame(x, y)

knot.counts <- 1:10  # Test from 2 to 10 knots

# Create a data frame to store the results
cv.results.ns <- data.frame(knots = knot.counts, cv.error.ns = NA) #setup for natural splines
cv.results.cs <- data.frame(knots = knot.counts, cv.error.cs = NA) #set up for cubic splines

# Perform cross-validation for each number of knots
for (i in 1:length(knot.counts)) {
  num.knots <- knot.counts[i]

  # Define the cross-validation folds
  folds <- createFolds(data$y, k = 10) # 10-fold CV

  # Store the errors for each fold
  fold.errors.ns <- numeric(length(folds))
  fold.errors.cs <- numeric(length(folds))

  # Perform cross-validation
  for (j in 1:length(folds)) {
    train.data <- data.test[-folds[[j]], ]
    test.data <- data.test[folds[[j]], ]

    # Fit the natural spline model
    spline.model.ns <- lm(y ~ ns(x, knots = quantile(train.data$x, seq(0, 1, length.out = num.knots + 2))[-c(1, num.knots+2)]), data = train.data) #the knots are the quantiles, but not the first or last (i.e., not the min or max)

     # Fit the cubic spline model using the b-splines bs() function
    spline.model.cs <- lm(y ~ bs(x, knots = quantile(train.data$x, seq(0, 1, length.out = num.knots + 2))[-c(1, num.knots+2)]), data = train.data) #the knots are the quantiles, but not the first or last (i.e., not the min or max)

    # Make predictions on the test fold
    predictions.ns <- predict(spline.model.ns, newdata = test.data)
    predictions.cs <- predict(spline.model.cs, newdata = test.data)

    # Calculate the mean squared error (MSE)
    fold.errors.ns[j] <- mean((test.data$y - predictions.ns)^2)
    fold.errors.cs[j] <- mean((test.data$y - predictions.cs)^2)
  }

  # Store the average MSE
  cv.results.ns$cv.error.ns[i] <- mean(fold.errors.ns)
  cv.results.cs$cv.error.cs[i] <- mean(fold.errors.cs)
}

# Find the optimal number of knots
optimal.knots.ns <- cv.results.ns$knots[which.min(cv.results.ns$cv.error.ns)]
cat("Optimal number of knots for natural spline", optimal.knots.ns, "\n")
optimal.knots.cs <- cv.results.cs$knots[which.min(cv.results.cs$cv.error.cs)]
cat("Optimal number of knots for cubic spline", optimal.knots.cs, "\n")

# Plot the cross-validation results
ggplot(cv.results.ns, aes(x = knots, y = cv.error.ns)) +
  geom_line(col="red") +
  geom_point(col="red") +
  geom_line(data=cv.results.cs, aes(x=knots, y=cv.error.cs), col="blue") +
  geom_point(data=cv.results.cs, aes(x=knots, y=cv.error.cs), col="blue") +
  labs(title = "Cross-Validation for Knot Selection (red is ns) (blue is cs)", x = "Number of Knots", y = "Cross-Validation Error (MSE)") +
  theme_minimal()

# Fit the final models with the optimal number of knots
final.spline.model.ns <- lm(y ~ ns(x, knots = quantile(data.test$x, seq(0, 1, length.out = optimal.knots.ns + 2))[-c(1, optimal.knots.ns+2)]), data = data.test)

final.spline.model.cs <- lm(y ~ bs(x, knots = quantile(data.test$x, seq(0, 1, length.out = optimal.knots.cs + 2))[-c(1, optimal.knots.cs+2)]), data = data.test)


# Visualize the final models
ggplot(data.test, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(y = predict(final.spline.model.ns)), color = "red") +
  geom_line(aes(y = predict(final.spline.model.cs)), color = "blue") +

  labs(title = paste("Natural (red) and Cubic (blue) Spline with", optimal.knots.10x, "Knots"), x = "X", y = "Y") +
  theme_minimal()
```

Just two knots actually seems optimal. Those knots would be located at:

```{r}
quantile(x,seq(0,1,length.out=4))# 7.522 and 15.044
```

But still, what is the test MSE for a model with 8, 6, or 2 equally spaced knots compared to knots at x=11 and x=16?

First, let's generate a new validation set of 100 x's using a different seed.

```{r}
set.seed(228)

x1 <- seq(0, 4 * pi, length.out = 100)
y1 <- sin(x1) + rnorm(100, 0, 0.3)
x2 <- seq(0, 10, length.out = 100)
y2 <- ifelse(x2 < 3, 2 * x2, ifelse(x2 < 7, -x2 + 9, 0.5 * x2 + 2)) + rnorm(100, 0, 0.5)

x <- x1+x2
y <- y1+y2
valid.data <- data.frame(x,y)
```

Let's fit the natural spline models (with original data and 10x data) with knots at x=11 and x=16.

```{r}
spline.model.2knots <- lm(y ~ ns(x, knots = c(11,16)), data = data)
spline.model.2knots.10x <- lm(y ~ ns(x, knots = c(11,16)), data = data.10x)
ggplot(valid.data, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(y = predict(spline.model.2knots.10x, newdata=valid.data)), color = "blue") +
  geom_line(aes(y = predict(spline.model.2knots, newdata=valid.data)), color = "red") +
  labs(title = "Natural Spline with 2 pre-specified knots (blue is 10x data)", x = "X", y = "Y") +
  theme_minimal()
```

